# Feature ID Conversion {#id-conversion}

This section shows how to convert from one feature ID (UniProt accessions, gene symbols, etc.) to another using different packages.

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```


## Conversion with `biomaRt`

```{r}
## Install missing packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
bio_pkgs <- c("biomaRt", "Biostrings")
for (pkg_i in bio_pkgs) {
  if (!require(pkg_i, quietly = T, character.only = T))
    BiocManager::install(pkg_i)
}
if (!require("remotes", quietly = T)) install.packages("remotes")
if (!require("MSnID", quietly = T, character.only = T))
  remotes::install_github("PNNL-Comp-Mass-Spec/MSnID@pnnl-master")
## ------------------------
library(biomaRt) # ID conversion
library(Biostrings) # read FASTA files
library(MSnID) # parse_FASTA_names
```

The first steps are to determine which mart and dataset to use. `listMarts` will show the available marts. The first 6 rows of the available datasets (provided by `listDatasets(mart)`) are also shown. (Use `View`, rather than `head`, to search for the desired database.)

```{r}
# Create mart
listMarts() # determine biomart for useMart
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL")
head(listDatasets(mart)) # determine dataset for useMart
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
                dataset = "rnorvegicus_gene_ensembl")
```

Next, we determine which attributes to select for the conversion table with `listAttributes(mart)`. From that table, we select "refseq_peptide" and "external_gene_name".

```{r}
# Create conversion table
head(listAttributes(mart)) # determine attributes for getBM
conv_tbl1 <- getBM(attributes = c("refseq_peptide", "external_gene_name"),
                   mart = mart)
head(conv_tbl1, 10)
```

This table has a lot of blank entries that need to be removed.


## Conversion with `AnnotationHub`

```{r}
## Install missing packages
if (!require("remotes", quietly = T)) install.packages("remotes")
if (!require("MSnID", quietly = T))
  remotes::install_github("PNNL-Comp-Mass-Spec/MSnID@pnnl-master")
```

`MSnID` has a function called `fetch_conversion_table` that leverages `AnnotationHub` to create a conversion table.

```{r}
# Create conversion table with MSnID::fetch_conversion_table
conv_tbl2 <- fetch_conversion_table(
  organism_name = "Rattus norvegicus", from = "REFSEQ", to = "SYMBOL"
)
head(conv_tbl2, 10)
```


## Conversion Using FASTA Headers

If specifically converting to gene symbols, it is recommended to use the information in the headers of the FASTA file that was used for the database search. The gene symbol is always given by `GN=...`, so we can use a regular expression to extract it. For UniProt FASTA files, there is a function in `MSnID` called `parse_FASTA_names` that will extract the components of the FASTA headers and create a `data.frame`.

```{r}
## Read FASTA file
fst_path <- system.file("extdata/uniprot_rat_small.fasta.gz",
                        package = "MSnID")
conv_tbl3 <- parse_FASTA_names(path_to_FASTA = fst_path)
head(conv_tbl3)
```

