
This has been removed from the pathway_analysis.Rmd file to make it less complicated. I am keeping it just in case I want to add some of it back.

```{r cp-table-format, include=FALSE}
# Function to format the result table of an enrichResult object
format_cp_ora <- function(x, caption, link, n = 10) {
  head(x@result, n) %>%
    # Reduce the number of gene IDs displayed.
    # Select the first three genes from each term and 
    # replace the rest with an ellipsis.
    # group_by(Cluster) %>% 
    # slice_min(order_by = p.adjust, n = 5, with_ties = FALSE) %>% 
    # ungroup() %>% 
    mutate(
      geneID = lapply(strsplit(geneID, "/"), function(genes) {
        paste0(paste(genes[1:min(3, length(genes))], collapse = "/"), 
               ifelse(length(genes) > 3, "...", ""))
      }),
      ID = cell_spec(
        ID, link = sprintf(link, ID))
    ) %>%
    # Nice table.
    kable(digits = 4, row.names = FALSE, caption = caption,
          escape = FALSE) %>% 
    kable_styling(full_width = FALSE, font_size = 12,
                  bootstrap_options = c("hover", "condensed")) %>% 
    # column_spec(2, width_max = "15em") %>% 
    scroll_box(height = "20em")
}
```

#### Gene Ontology {#ora-go}

We will first use the clusterProfiler package to test which biological processes are over-represented in each cluster. For this example, we will only consider gene sets of size 15 to 300. In order to test either molecular functions, cellular components, or all three ontologies at once, set `ont` to `"MF"`, `"CC"`, or `"ALL"`, respectively.

```{r cp-ora-go}
## GO BP ORA with clusterProfiler
# This takes a while
cp_ora_go <- compareCluster(
  geneClusters = gcUnique, 
  fun = "enrichGO", # ORA function to apply to each cluster
  # Arguments below are passed to enrichGO
  OrgDb = "org.Hs.eg.db", 
  keyType = "ENTREZID", 
  ont = "BP", # BP, CC, MF, or ALL for all ontologies
  pvalueCutoff = 0.05,
  qvalueCutoff = 1, # Do not filter by q-value
  pAdjustMethod = "BH", # p-values are adjusted within clusters
  universe = universe, # all genes
  minGSSize = 15, 
  maxGSSize = 300, 
  readable = TRUE # Convert Entrez ID to gene symbol
)
```

```{r cp-ora-go-table, echo=FALSE, fig.alt="Top significantly over-represented biological processes from enrichGO output."}
# Top 10 terms
format_cp_ora(cp_ora_go, 
              caption = paste("Top significantly over-represented biological",
                              "processes from enrichGO output."),
              link = "https://www.ebi.ac.uk/QuickGO/term/%s")
```

</br>

There were `r length(cp_ora_go@geneSets)` biological processes that were tested. Of these, `r nrow(cp_ora_go@result)` passed the size filter, and only `r sum(cp_ora_go@result$p.adjust < 0.05)` were significantly over-represented after multiple testing correction. The top 10 are shown in Table \@ref(tab:cp-ora-go-table).

**Remove Redundant GO Terms**

By default, if two terms have a semantic similarity above 0.7, `clusterProfiler::simplify` retains the more significantly over-represented term.

```{r simplify-cp-ora-bp}
# Remove redundant GO terms with simplify
cp_ora_go_sim <- simplify(cp_ora_go)
```

```{r simplify-cp-ora-bp-table, echo=FALSE, fig.alt="Simplified top significantly over-represented biological processes from enrichGO output."}
format_cp_ora(cp_ora_go_sim, 
              caption = paste("Simplified top significantly over-represented", 
                              "biological processes from enrichGO output."),
              link = "https://www.ebi.ac.uk/QuickGO/term/%s")
```

</br>

Notice that <a href="https://www.ebi.ac.uk/QuickGO/term/GO:0009060" title = "aerobic respiration">GO:0009060</a> is not present in Table \@ref(tab:simplify-cp-ora-bp-table) anymore, since it was highly similar to <a href="https://www.ebi.ac.uk/QuickGO/term/GO:0006119" title = "oxidative phosphorylation">GO:0006119</a>, and the latter was more significant. After removing redundant GO terms, `r sum(cp_ora_go_sim@result$p.adjust < 0.05)` pass the significance threshold.

Instead of retaining terms based on adjusted p-value, we could ignore significance and instead retain either parent or child terms (`select_fun = min` and `select_fun = max`, respectively). The code to do so is provided <a href="https://github.com/YuLab-SMU/clusterProfiler/issues/372" title = "New feature: simplify to parent GO terms #372">here</a>.

Another alternative would be to use a gene ID to <a href="http://geneontology.org/docs/go-subset-guide/" title = "Guide to GO subsets">GO slim</a> ID conversion table. See the Semantic Similarity subsection of Section \@ref(annotation-databases) for more details.

Now, we will perform GO ORA with the GOstats package. We begin by creating an object of class `GOHyperGParams` and passing it to `hyperGTest` for calculation of Hypergeometric p-values. We do not filter by p-value, since adjusted p-values are not provided and we must calculate them ourselves.

```{r gostats-go-ora, message=FALSE}
## GO BP ORA with GOstats
hyperg_ora_go <- new(
  Class = "GOHyperGParams",
  ontology = "BP",
  geneIds = sig_genes,
  universeGeneIds = universe,
  annotation = "org.Hs.eg.db",
  pvalueCutoff = 1, # Do not filter by p-value
  testDirection = "over",
  conditional = FALSE
) %>%
  hyperGTest() # Hypergeometric testing
```

Now, we need to filter by gene set size (limit 20 to 500), adjust the p-values, and filter based on the adjusted p-values.

```{r hyperg-go-ora-res}
res_go_ora <- summary(hyperg_ora_go) %>%
  # Filter by size
  filter(Size >= 20, Size <= 500) %>%
  # Adjust p-values
  mutate(p.adjust = p.adjust(Pvalue, method = "BH")) %>%
  # Filter and sort by adjusted p-values
  arrange(p.adjust) %>% 
  filter(p.adjust < 0.05)
```

```{r include=FALSE}
n1 <- summary(hyperg_ora_go)
n2 <- filter(n1, Size >= 20, Size <= 500) %>% nrow()
```

```{r cp-go-ora-table, echo=FALSE, fig.alt="Over-represented biological processes obtained using the GOstats package."}
res_go_ora[, c("GOBPID", "Term", "OddsRatio", "ExpCount", 
               "Count", "Size", "Pvalue", "p.adjust")] %>%
  head(10) %>%
  # Add links to terms
  mutate(GOBPID = cell_spec(
    GOBPID, 
    link = sprintf("https://www.ebi.ac.uk/QuickGO/term/%s", GOBPID))) %>%
  kable(digits = 4, row.names = FALSE, escape = FALSE,
        caption = paste("Over-represented biological processes", 
                        "obtained using the GOstats package.")) %>% 
  kable_styling(full_width = FALSE, font_size = 12,
                bootstrap_options = c("hover", "condensed")) %>% 
  scroll_box(height = "20em")
```

</br>

There were `r nrow(n1)` biological processes that were tested. Of these, `r n2` passed the size filter, and only `r nrow(res_go_ora)` were significantly over-represented after multiple testing correction. The top 10 are shown in Table \@ref(tab:cp-go-ora-table). The total number of biological processes that were tested is 1 less than the number reported from the `enrichGO` results. This is because `enrichGO` includes the "biological_process" root <a href="https://www.ebi.ac.uk/QuickGO/term/GO:0008150" title = "biological process">GO:0008150</a>.

**Remove Redundant GO Terms**

While there is no function akin to `clusterProfiler::simplify` to process results, there is an argument called `conditional`. Setting `conditional` to `TRUE` tells `hyperGTest` to use the term-term relationships when calculating p-values [@falcon_how_2021]. This approach is actually more akin to Modular Enrichment Analysis (MEA) than Over-Representation Analysis (also called Singular Enrichment Analysis; @huang_bioinformatics_2009).

```{r gostats-go-ora-simple, message=FALSE}
## Conditional GO BP ORA with GOstats
hyperg_ora_go_sim <- new(
  Class = "GOHyperGParams",
  ontology = "BP",
  geneIds = sig_genes,
  universeGeneIds = universe,
  annotation = "org.Hs.eg.db",
  pvalueCutoff = 1, # Do not filter by p-value
  testDirection = "over",
  conditional = TRUE # Use structure of GO graph
) %>%
  hyperGTest() # Hypergeometric testing
```

We will process the results the same as before and compare.

```{r hyperg-go-ora-res-simple, echo=FALSE}
res_go_ora_sim <- summary(hyperg_ora_go_sim) %>%
  # Filter by size
  filter(Size >= 20, Size <= 500) %>%
  # Adjust p-values
  mutate(p.adjust = p.adjust(Pvalue, method = "BH")) %>%
  # Filter and sort by adjusted p-values
  arrange(p.adjust) %>% 
  filter(p.adjust < 0.05)
```

```{r include=FALSE}
n3 <- summary(hyperg_ora_go_sim)
n4 <- filter(n3, Size >= 20, Size <= 500) %>% nrow()
```

```{r cp-go-ora-simple-table, echo=FALSE, fig.alt="Simplified over-represented biological processes obtained using the GOstats package."}
res_go_ora_sim[, c("GOBPID", "Term", "OddsRatio", "ExpCount", 
                   "Count", "Size", "Pvalue", "p.adjust")] %>%
  head(10) %>%
  # Add links to terms
  mutate(GOBPID = cell_spec(
    GOBPID, 
    link = sprintf("https://www.ebi.ac.uk/QuickGO/term/%s", GOBPID))) %>%
  kable(digits = 4, row.names = FALSE, escape = FALSE,
        caption = paste("Simplified over-represented biological processes", 
                        "obtained using the GOstats package.")) %>% 
  kable_styling(full_width = FALSE, font_size = 12,
                bootstrap_options = c("hover", "condensed")) %>% 
  scroll_box(height = "20em")
```

</br>

Now, only `r nrow(n3)` terms showed up in the summary. Of these, `r n4` passed the size filter, and `r nrow(res_go_ora_sim)` were significantly over-represented after multiple testing correction. The top 10 are shown in Table \@ref(tab:cp-go-ora-simple-table). The major difference between these results and the ones from `enrichGO` after using `simplify` is that p-value adjustment is not affected by `simplify`, as redundant GO terms are removed after. The conditional method is actually more akin to performing modular enrichment analysis (MEA) than ORA (also called singular enrichment analysis)


#### Reactome {#ora-reactome}

GOstats does not have a dedicated class to test for Reactome pathway over-representation, but we can use ReactomePA. For this example, we will only consider pathways of size 20 to 500. `enrichPathway` only accepts Entrez gene IDs as input. In cases where the Entrez ID is not readily available, we must convert to them. Alternatively, we could use the custom ORA function `enricher`, which allows us to use any ID type (use demonstrated in Pfam example).

```{r cp-ora-react}
## Reactome ORA with ReactomePA
cp_ora_reactome <- enrichPathway(
  gene = sig_genes, 
  organism = "human", 
  pvalueCutoff = 0.05, 
  pAdjustMethod = "BH", 
  qvalueCutoff = 1, # Do not filter by q-value
  universe = universe, 
  minGSSize = 20, 
  maxGSSize = 500,
  readable = TRUE # Convert Entrez ID to gene symbol
)
```

```{r cp-ora-react-table, echo=FALSE, fig.alt="Top significantly over-represented Reactome pathways from enrichPathway output."}
format_cp_ora(cp_ora_reactome, 
              caption = paste("Top significantly over-represented", 
                              "Reactome pathways from enrichPathway output."),
              link = "https://reactome.org/content/detail/%s")
```

</br>

There were `r length(cp_ora_reactome@geneSets)` Reactome pathways that were tested. Of these, `r nrow(cp_ora_reactome@result)` passed the size filter, and `r sum(cp_ora_reactome@result$p.adjust < 0.05)` were significantly over-represented after multiple testing correction. The top 10 are shown in Table \@ref(tab:cp-ora-react-table).

<!---
#### KEGG {#ora-kegg}

```{r cp-ora-kegg, eval=FALSE}
## KEGG ORA with clusterProfiler
cp_ora_kegg <- enrichKEGG(
gene = sig_genes,
organism = "rno", # Rattus norvegicus
keyType = "ncbi-proteinid",
pvalueCutoff = 0.05,
pAdjustMethod = "BH",
universe = universe,
minGSSize = 20,
maxGSSize = 500,
qvalueCutoff = 1,
use_internal_data = FALSE
)
```
--->

#### Pfam {#ora-pfam}

Now, we will perform Pfam ORA with the GOstats package. We begin by constructing a new object of class `PFAMHyperGParams` and then passing it to `hyperGTest` for calculation of Hypergeometric p-values. We do not filter by p-value, since adjusted p-values are not provided and we must calculate them ourselves.

```{r gostats-pfam-ora, message=FALSE}
## Pfam ORA with GOstats
hyperg_pfam_ora <- new(
  Class = "PFAMHyperGParams",
  geneIds = sig_genes,
  universeGeneIds = universe,
  annotation = "org.Hs.eg.db",
  pvalueCutoff = 1,
  testDirection = "over"
) %>%
  hyperGTest() # Hypergeometric testing
```

Now, we need to filter by gene set size (limit 15 to 500) and adjust the p-values. Normally, we would also filter based on the adjusted p-values, but we will skip this step so that we can compare the output with that of `clusterProfiler::enricher`.

```{r include=FALSE}
n5 <- summary(hyperg_pfam_ora)
n6 <- filter(n5, Size >= 20, Size <= 500) %>% nrow()
```

```{r res-pfam-ora}
res_pfam_ora <- summary(hyperg_pfam_ora) %>%
  # Filter by Pfam entry size
  filter(Size >= 15, Size <= 500) %>%
  # Adjust p-values
  mutate(p.adjust = p.adjust(Pvalue, method = "BH")) %>%
  # Sort by adjusted p-values
  arrange(p.adjust)
```

<!---

The last step is to add the human-readable Pfam descriptions. We will use the `TERM2NAME` data frame from before and just change the column names for easy joining.

```{r process-pfam-gostats, message=FALSE, eval=FALSE}
# Pfam entries to descriptions
colnames(TERM2NAME) <- c("PFAMID", "Description")
res_pfam_ora <- left_join(res_pfam_ora, TERM2NAME)
```

--->

```{r gostats-pfam-ora-table, echo=FALSE, fig.alt="Top Pfam entries obtained using the GOstats package."}
res_pfam_ora[, c("PFAMID", "OddsRatio", "ExpCount", 
                 "Count", "Size", "Pvalue", "p.adjust")] %>%
  head(10) %>% 
  mutate(
    PFAMID = cell_spec(
      PFAMID, 
      link = sprintf("http://pfam.xfam.org/family/%s", PFAMID))
  ) %>%
  kable(digits = 4, row.names = FALSE, escape = FALSE,
        caption = "Top Pfam entries obtained using the GOstats package.") %>% 
  kable_styling(full_width = FALSE, font_size = 12,
                bootstrap_options = c("hover", "condensed")) %>% 
  scroll_box(height = "20em")
```

</br>

`r nrow(n5)` Pfam entries were tested. Of these, `r n6` passed the size filter, and none were significantly over-represented after multiple testing correction. The top 10 are shown in Table \@ref(tab:gostats-pfam-ora-table).


#### Other Databases {#ora-other}

For databases without a dedicated ORA function, we use `clusterProfiler::enricher`, which is a general ORA function that can be used with any term to gene conversion table. This function is also much faster than the dedicated `enrichGO` and `enrichPathway` functions (since the conversion table does not need to be fetched), so it may be preferred if running many separate over-representation analyses.

**NOTE: The term to gene conversion table must have exactly two columns: term and gene. They must also be in that exact order!**

For this example, we will perform Pfam ORA. The conversion table can be generated with the biomaRt package as follows.

```{r eval=FALSE}
library(biomaRt)

mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "hsapiens_gene_ensembl")
View(listAttributes(mart)) # Determine which columns to return

TERM2GENE <- getBM(filters = "entrezgene_id",
                   attributes = c("entrezgene_id",
                                  "pfam"),
                   values = universe,
                   mart = mart) %>%
  mutate_all(as.character) %>% 
  # !!! These specific column names must be in this exact order
  dplyr::select(term = pfam, gene = entrezgene_id) %>% 
  dplyr::filter(term != "") # remove empty terms
```

```{r eval=FALSE, include=FALSE}
# Save table
save(TERM2GENE, file = "data/pfam_term2gene.RData", compress = TRUE)
```

In addition to the required term to gene conversion table, we can supply an optional term to name table that maps Pfam IDs to human-readable descriptions. While org.Hs.eg.db does not provide Pfam descriptions, the <a href="https://doi.org/doi:10.18129/B9.bioc.PFAM.db" title = "PFAM.db: A set of protein ID mappings for PFAM">PFAM.db</a> annotation data package has a Bimap with this information. Like the `TERM2GENE` table, name and order of columns matters.

```{r eval=FALSE}
library(PFAM.db)

# Pfam IDs to descriptions
TERM2NAME <- as.data.frame.Bimap(PFAMDE) %>% 
  # !!! These specific column names must be used in this exact order
  setNames(c("term", "name"))
```

```{r include=FALSE, eval=FALSE}
# Save table
save(TERM2NAME, file = "data/pfam_term2name.RData", compress = TRUE)
```

Now that we have the two conversion tables, we will perform ORA and limit the output to Pfam entries of size 15 to 500.

```{r echo=FALSE, results="hide"}
load("data/pfam_term2gene.RData")
load("data/pfam_term2name.RData")
```


```{r cp-ora-pfam}
# Pfam ORA with clusterProfiler
cp_ora_pfam <- enricher(
  gene = sig_genes,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  universe = universe,
  minGSSize = 15,
  maxGSSize = 500,
  qvalueCutoff = 1,
  TERM2GENE = TERM2GENE,
  TERM2NAME = TERM2NAME
)
```

```{r cp-ora-pfam-table, echo=FALSE, fig.alt="Top Pfam entries from enricher output."}
format_cp_ora(
  cp_ora_pfam, 
  caption = "Top Pfam entries from enricher output.",
  link = "http://pfam.xfam.org/family/%s", 
  n = 10
)
```

</br>

There were `r length(cp_ora_pfam@geneSets)` Pfam entries that were tested. Of these, `r nrow(cp_ora_pfam@result)` passed the size filter, and `r sum(cp_ora_pfam@result$p.adjust < 0.05)` were significantly over-represented after multiple testing correction. While none were significant, the top 10 are shown in Table \@ref(tab:cp-ora-pfam-table).

Notice that this table and Table \@ref(tab:gostats-pfam-ora-table) are different. For one, the sizes of the entries are not always the same. For example, <a href="http://pfam.xfam.org/family/PF01391" title = "Collagen triple helix repeat (20 copies)">PF01391</a> has size 31 according to the mapping file created from the `biomaRt` package, but only size 30 according to `org.Hs.eg.db` (used by GOstats). This has to do with how up-to-date the information is and leads to differences in the number of total Pfam entries that remain after filtering by size and, subsequently, differences in the adjusted p-values.

<!---
First, we will perform Pfam ORA with the GOstats package using the org.Hs.eg.db human annotation database; however, it is a good idea to check the number of Pfam entries that are provided by this database before doing so. Ideally, it should be close to the number shown at the top of the <a href="http://pfam.xfam.org">main Pfam page</a>. We can accomplish this with functions in AnnotationDbi.

```{r entrez-to-pfam, message=FALSE}
# Map from Entrez gene to Pfam ID
ENTREZ2PFAM <- AnnotationDbi::select(
org.Hs.eg.db, keys = keys(org.Hs.eg.db, keytype = "ENTREZID"),
columns = c("ENTREZID", "PFAM"), keytype = "ENTREZID"
)

# Nice table
kable(head(ENTREZ2PFAM)) %>% 
kable_styling(full_width = FALSE, font_size = 12)
```

In this data frame, there are only `r length(unique(ENTREZ2PFAM$PFAM))` Pfam entries. This is not even close to the most up-to-date number. We will continue for the sake of demonstration, but the results are unlikely to be reliable. 
--->


<!---
------------------------------------------------------------------------------
--->












```{r include=FALSE}
# Do not run below chunks
knitr::opts_chunk$set(eval=FALSE)
```


<!---
Suppose we have a table of differential analysis results with columns for the feature being tested, the comparison being done, and the resulting log$_2$ fold-changes, t-statistics, p-values, and adjusted p-values. 

Gene Set Enrichment Analysis (GSEA) determines whether genes in a predefined set are primarily up- or down-regulated. 

The input of GSEA is a uniquely-named vector of ranking metric values. The names are usually Entrez gene IDs, but they can be any type of ID, so long as there is a table that maps all or most of these IDs to predefined sets (biological processes, molecular functions, Reactome pathways, etc.). The ranking metric must satisfy two criteria: the sign indicates the direction of change, and the magnitude indicates importance (usually some indicator of statistical significance). Two of the more commonly-used ranking metrics are t-statistics and the $-log_{10}(\text{p-value}) * sign(logFC)$.

Once we have this named vector, the procedure is as follows:

1) Sort the vector of $N$ genes in descending order. Call this $L$.
2) For an *a priori* defined gene set $S$, calculate an enrichment score. This is done by "walking down... $L$, increasing a running-sum statistic when we encounter a gene in $S$ and decreasing it when we encounter genes not in $S$." The maximum deviation from zero of this running-sum statistic is the enrichment score for $S$, denoted by $ES(S)$. If the genes in $S$ are randomly distributed throughout $L$, $ES(S)$ will be relatively small; however, if they are not randomly distributed (i.e. primarily located near either end of $L$), then $ES(S)$ will be relatively large [@mootha_pgc-1-responsive_2003; @subramanian_gene_2005].
3) 

--->
GSEA employs a "no-cutoff" strategy that utilizes some experimental value (such as fold change, the moderated t-statistic, or Z-Score) to rank a list of $N$ genes in descending order. Using this ranked list $L$, the values of the ranking metric, and an *a priori* defined gene set $S$, we can calculate an enrichment score. This is done by "walking down the list $L$, increasing a running-sum statistic when we encounter a gene in $S$ and decreasing it when we encounter genes not in $S$." The maximum deviation from zero of this running-sum statistic is the enrichment score for $S$, denoted by $ES(S)$. If the genes in $S$ are randomly distributed throughout $L$, $ES(S)$ will be relatively small; however, if they are not randomly distributed (i.e. primarily located near either end of $L$), then $ES(S)$ will be relatively large [@mootha_pgc-1-responsive_2003; @subramanian_gene_2005].

To assess the statistical significance of each $ES(S)$, we permute the phenotype labels in the expression data and recalculate the ranking metric and subsequent enrichment score $ES_{\text{NULL}}$. Alternatively, we could permute the genes: for a given $S$ of size $k$, we randomly select $k$ genes from the list $L$ and calculate an enrichment score $ES_{\text{NULL}}$. This is the method implemented by <a href="https://doi.org/doi:10.18129/B9.bioc.fgsea" title = "fgsea: Fast Gene Set Enrichment Analysis">fgsea</a> and, consequently, <a href="https://doi.org/doi:10.18129/B9.bioc.clusterProfiler" title = "clusterProfiler: A universal enrichment tool for interpreting omics data">clusterProfiler</a>, though it is "not strictly accurate" because it does not preserve the correlations between genes [@subramanian_gene_2005]. Regardless of the permutation approach, it is repeated one thousand to tens of thousands of times to produce an empirical null distribution. Then, we define the nominal p-value as the proportion of $ES_{\text{NULL}}$ that are more extreme than $ES(S)$. Once a p-value has been obtained for every gene set, we apply an appropriate adjustment for multiple hypothesis testing. The phenotype randomization approach is summarized in <a href="https://www.nature.com/articles/ng1180/figures/1" title = "Figure 1: Schematic overview of GSEA.">Figure 1: Schematic overview of GSEA</a> in the paper by @mootha_pgc-1-responsive_2003. The mathematical details of GSEA are provided below.

Let $x = \{x_1, ..., x_N : x_i > x_j \ \forall i < j\}$ be the vector of ranking metric values. Also let $S$ be a particular gene set of size $k$ and $NS = \sum\limits_{i \in S}|x_i|$. We define the value of the running sum at gene $i$ as

<div class="math">
\[
r_i = \begin{cases}
0 & \text{if } i = 0, \\
r_{i-1} + \frac{1}{NS}|x_i| & \text{if } 1 \leq i \leq N \text{ and } i \in S, \\
r_{i-1} - \frac{1}{N-k} & \text{if } 1 \leq i \leq N \text{ and } i \not\in S
\end{cases}
\]
</div>

$ES(S)$ is the largest (in terms of absolute value) element of $r_N$. This notation is a slightly modified version of what is presented in the paper by @korotkevich_fast_2016.

**Important Considerations**
<ol>
<li>
GSEA is not influenced by an arbitrary cutoff for statistical significance. This is especially useful when only a few features pass this threshold after adjustment for multiple testing, and it also means that it does not depend on the choice of p-value adjustment.
</li>
<li>
All genes in an experiment are allowed to "contribute to the enrichment analysis in differing degrees" [@huang_bioinformatics_2009]. This addresses the point brought up at the beginning of Section \@ref(pathway-analysis) that biological changes may be the result of small changes in many related genes. GSEA allows for the detection of such changes.
</li>
<li>
The sign of the enrichment score tell us in which category a set of genes is primarily up- or down-regulated. For example, suppose we test the difference between the means of two groups for each gene using the <a href="https://doi.org/doi:10.18129/B9.bioc.limma" title = "limma: Linear Models for Microarray Data">limma</a> package. Since ranking metrics incorporate the direction of change, a positive enrichment score would mean that the majority of genes in that particular set are up-regulated in A relative to B/down-regulated in B relative to A.
</li>
<li>
The choice of ranking metric is important: `-log10(p-value) * sign(logFC)`, `-log10(p-value) * logFC`, t-statistics, and Z-Scores are just a few possible choices.
</li>
</ol>


### Examples

```{r warning=FALSE, message=FALSE}
## Install missing packages
cran_packages <- c("remotes", "dplyr", "BiocManager")
for (pkg_i in cran_packages) {
  if (!require(pkg_i, quietly = T, character.only = T))
    install.packages(pkg_i)
}
bio_packages <- c("org.Hs.eg.db", "clusterProfiler", "ReactomePA")
for (pkg_i in bio_packages) {
  if (!require(pkg_i, quietly = T, character.only = T))
    BiocManager::install(pkg_i)
}
git_packages <- c("MSnID@pnnl-master", "MSnSet.utils")
for (pkg_i in git_packages) {
  if (!require(sub("@.*", "", pkg_i), quietly = T, character.only = T))
    remotes::install_github(file.path("PNNL-Comp-Mass-Spec", pkg_i))
}
## ------------------------
library(MSnID) # fetch_conversion_table
library(MSnSet.utils) # oca.set data
library(org.Hs.eg.db) # Human database package
library(clusterProfiler) # GO and custom GSEA
library(ReactomePA) # Reactome GSEA
library(dplyr)

data(cptac_oca) # data
m <- oca.set
```

The input for the GSEA functions is a named vector of ranking metric values (t-statistics, F-statistics, -log10(p-value) * sign(logFC), etc.) where the names are typically unique Entrez gene IDs. We need to go from an MSnSet to differential analysis results to this named vector. First, we will get the Entrez IDs for each RefSeq and add this as a column in `fData`.

```{r message=FALSE}
# Get RefSeq to Entrez ID conversion table
conv_tbl <- fetch_conversion_table(organism_name = "Homo sapiens", 
                                   from = "REFSEQ", 
                                   to = "ENTREZID") %>% 
  setNames(c("RefSeq_no_iso", "entrez_gene")) # rename columns
head(conv_tbl)
```


```{r}
# Add RefSeq_no_iso and entrez_gene columns to fData(m)
temp <- fData(m) %>% 
  # Remove isoform number from RefSeq
  mutate(RefSeq_no_iso = gsub("\\.\\d+", "", RefSeq)) %>% 
  # Add entrez_gene column
  left_join(conv_tbl, by = "RefSeq_no_iso")
rownames(temp) <- temp$RefSeq # Set rownames to RefSeq column

fData(m) <- temp # overwrite fData
head(fData(m))
```

Not every RefSeq mapped to an Entrez gene. We will not be able to use these features, and the code below shows how many were mapped/not mapped.

```{r}
table(!is.na(fData(m)$entrez_gene))
```

109 proteins (~1.3% in table below) were not mapped to any gene. We will remove those rows.

```{r}
100 * prop.table(table(!is.na(fData(m)$entrez_gene)))
```

Now that we have an Entrez ID column in the `fData`, we can move on to the next steps. I have combined all steps below so that the code is easier to copy. First, we need a table of differential analysis results. From there, we subset to rows without any missing values. This removes proteins that did not map to a gene, as well as proteins with no test statistics or p-values. Then, we create a column for the ranking metric. We will use $-log10(\text{p-value}) \cdot sign(\text{log}_2 \text{ fold-change})$. We could have also used the moderated t-statistic, which is similar (see Figure \@ref(fig:rank-metric-comp)). Now, we need to make sure that there is only one value per gene. We do so by calculating the average ranking metric for each gene. The last step is to sort from high to low by ranking metric and convert to a named vector. We can see the first and last 6 entries of this named vector below.

```{r}
## Named vector for GSEA
# Start with differential analysis results
gsea_input <- limma_gen(m, model.str = "~ PLATINUM.STATUS", 
                        coef.str = "PLATINUM.STATUS") %>% 
  mutate(RefSeq = rownames(.)) %>% # Create RefSeq column
  left_join(fData(m), by = "RefSeq") %>% # Add entrez_gene column
  # Remove rows missing entrez_gene or P.Value
  filter(!if_any(c(entrez_gene, P.Value), is.na)) %>%
  # Create GSEA ranking metric column: signed -log10 p-value
  mutate(ranking_metric = -log10(P.Value) * sign(logFC)) %>% 
  # Average ranking metric for each gene
  group_by(entrez_gene) %>% 
  summarise(ranking_metric = mean(ranking_metric)) %>% 
  # Sort from high to low by ranking metric
  arrange(-ranking_metric) %>%
  # Convert to named vector - first column is names, second is values
  tibble::deframe()

head(gsea_input)
```

```{r rank-metric-comp, echo=FALSE, fig.cap="Comparison of two common GSEA ranking metrics."}
gsea_input2 <- limma_gen(m, model.str = "~ PLATINUM.STATUS", 
                         coef.str = "PLATINUM.STATUS") %>% 
  mutate(RefSeq = rownames(.)) %>%
  left_join(fData(m), by = "RefSeq") %>%
  filter(!is.na(entrez_gene), !is.na(P.Value)) %>%
  mutate(ranking_metric = t) %>% 
  group_by(entrez_gene) %>% 
  summarise(ranking_metric = mean(ranking_metric)) %>% 
  arrange(-ranking_metric) %>%
  tibble::deframe()

input_df <- data.frame(rank_metric = c(gsea_input, gsea_input2),
                       type = rep(c("signed_pval", "t_stat"), 
                                  each = length(gsea_input)))

library(ggplot2)

ggplot(input_df) +
  geom_density(aes(x = rank_metric, fill = type), color = NA, alpha = 0.5) +
  scale_x_reverse() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_classic(base_size = 14) +
  labs(x = "Value of Ranking Metric", y = "Density") +
  scale_fill_manual(
    name = NULL,
    values = c("#DB8124", "#247EDB"), 
    labels = c(expression(paste("-log"[10],"(p-value) * sign(logFC)")), 
               "Moderated t-statistic")
  ) +
  theme(legend.position = "bottom", 
        legend.direction = "horizontal")
```

Now that we have the input vector, we can move on to the examples.

```{r include=FALSE}
format_cp_gsea <- function(x, caption, link, n = 10) {
  head(x@result, n) %>%
    # Reduce the number of gene IDs displayed.
    # Select the first three genes from each term and 
    # replace the rest with an ellipsis.
    mutate(
      core_enrichment = lapply(strsplit(core_enrichment, "/"), function(genes) {
        paste0(paste(genes[1:min(3, length(genes))], collapse = "/"), 
               ifelse(length(genes) > 3, "...", ""))
      }),
      ID = cell_spec(
        ID, link = sprintf(link, ID))
    ) %>%
    # Nice table.
    kable(digits = 4, row.names = FALSE, caption = caption,
          escape = FALSE) %>% 
    kable_styling(full_width = FALSE, font_size = 12,
                  bootstrap_options = c("hover", "condensed")) %>% 
    # column_spec(2, width_max = "15em") %>% 
    scroll_box(height = "20em")
}
```


#### Gene Ontology

```{r}
# Biological Process GSEA
go_gsea <- gseGO(geneList = gsea_input, 
                 ont = "BP", 
                 eps = 0,
                 OrgDb = org.Hs.eg.db, 
                 nPermSimple = 1000) # may need to increase nPermSimple
# View(go_gsea@result) # uncomment to view results
```

```{r echo=FALSE}
format_cp_gsea(go_gsea, link = "https://www.ebi.ac.uk/QuickGO/term/%s", caption = " ")
```



#### Reactome

```{r}
# Reactome GSEA
react_gsea <- gsePathway(geneList = gsea_input,
                         organism = "human", 
                         eps = 0,
                         nPermSimple = 1000)
# View(react_gsea@result) # uncomment to view results
```


#### Other Databases {#gsea-other}

```{r eval=FALSE}
# GSEA with any database
other_gsea <- GSEA(geneList = gsea_input, 
                   eps = 0, 
                   TERM2GENE = TERM2GENE, 
                   TERM2NAME = TERM2NAME, 
                   nPermSimple = 1000)
# View(pfam_gsea@result) # uncomment to view results
```






